Toronto's Homeless: Understanding the Causes and Seeking Solutions

### 2023-04-20 outputs/paper/paper.qmd
final commit

---
title: "The Plight of Toronto's Homeless: Understanding the Causes and Seeking Solutions"
subtitle: "My subtitle if needed"
author: 
  - Will Davidson
thanks: "Code and data are available at: https://github.com/willdavidson01/Final-Paper/tree/main/outputs/paper."
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "People suffering from homelessness in the City of Toronto have become an accepted part of the urban landscape.  With the shortage of affordable housing, high rents, poor economy and rising mental health concerns, it’s hard to imagine a viable solution to this critical social determinant of health.  Using OpenDataToronto, this paper analyzes data and identifies trends in different populations, gender, and age, as well as the current capacity to house people either in an emergency or permanent housing. It also challenges current assumptions and draws attention to the fact that today’s data can provide solutions to tomorrow’s needs."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Despite economic growth and developments in the city, Toronto continues to work to help people experiencing homelessness, with the issue worsening over time. We've seen Toronto try and take action through different housing and recreation centers, but these attempts have fallen short of solving the problem. Individuals have struggled through extreme weather conditions and overcrowded shelters, with waitlists of up to ten years for subsidized housing. The supportive shelters can't keep up with the demand because of the intake of refugees, Toronto's condo boom, and Toronto's fight against drug and alcohol abuse. The complex issue at hand requires a complex solution from the city. Inadequate income, inaccessible housing, racism in employment, and housing markets contribute to homelessness. Regardless of socioeconomic status, fairness, inclusion, and opportunity need to happen to give each individual a chance to thrive. Everyone should be on the same playing field and given the same advantages, but this isn't the case, as some people are dealt the wrong hands and need help working through it. Homeless people are marginalized, and they're pushed to the side and deemed less important. The government needs to take urgent action to address the root causes of homelessness, as living on the streets puts individuals at risk of poor health and premature death. Should the government declare a state of emergency? Should more homeless shelters be built? What are the most significant issues the government faces when finding a solution? These critical questions need to be asked to move forward as one.

The data used in the report is obtained from OpenDataToronto. The dataset is called "Toronto Shelter System Flow," It provides information on who in Toronto is accessing overnight services between 2019 and 2023. I'm focused on the year the dataset is from, the number of homeless individuals, and how they're represented in the Toronto population. In section 2, a cleaned dataset is obtained and explained to provide further analysis. Visualizations in graphs and tables are presented to illustrate the correlation between year, homelessness, category, age, and race. By data visualization, we assume the number of homeless people over the past five years has risen before constructing the model. Section 3 presents a multiple linear regression model to justify the relationship between homelessness and year. The interpretation of the final model and reasoning is shown in the latter section. A discussion is carried out in section 5 on the implications of the findings, including causes and solutions. We also conclude with the following steps that need to be taken for a better future. 4

@@

# Data Collection

The raw data was obtained from Opendatatoronto within the "Toronto System Shelter Flow". The data for the System Flow is captured through an information management system used to operate overnight shelters, 24-hour repsite sites, and warming centers funded by the City of Toronto. The data is calculated on the last day of each month and it considers all people who accessed overnight shelters. The analysis of the data was done through the R programming language [@citeR] with the tidyverse [@citeTidyverse], dplyr [@citedplyr], and knitr (@citeknitr) packages being used to clean the data. ggplot2 [@citeggplot2] was used to produce all the figures in this paper. The Lubridate [@citelubridate] package was also used to reformat dates.

@@

```{r} 
#| label: fig-1
#| fig-cap: "Homeless Population Groups in Toronto From 2019 to 2023"
#| echo: false
#| warning: false
library(tidyverse)

ggplot(data = aux, aes(x = date, y = actively_homeless, color = population_group)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic() +
  labs(title = "Scatter Plot",
       x = "Date",
       y = "Actively Homeless")+
  scale_x_date(date_breaks = "12 months", date_labels = "%Y") 
```

Figure 1 represents a scatter plot that shows the trends of in Toronot's homelessness in the past five years. The Y-Axis is the number of current Torontoians who are actively homeless and the X-Axis are the twelve months in each of the past five years. The population groups are represented by individual data points.

```{r}
#| label: fig-2
#| fig-cap: "Homeless Population Based on Gender"
#| echo: false
#| warning: false
library(tidyverse)

ggplot(data = data_long, aes(x = date, y = value, color = Category)) + 
  geom_line(size = 2) +
  theme_classic() +
  labs(title = "Line Graph",
       x = "Date",
       y = "Actively Homeless")+
  scale_x_date(date_breaks = "12 months", date_labels = "%Y") 
```

Figure two are multiple line graphs that compare the number of actively homeless people based on gender over the past five years.

```{r}
#| label: fig-3
#| fig-cap: "Homeless Population by Age Group in Toronto"
#| echo: false
#| warning: false
library(tidyverse)
#3rd graph
aux = df %>% 
  select(date.mmm.yy., ageunder16, age16.24, age25.44, age45.64, age65over) %>% 
  filter(population_group == 'All Population' & #or chronic ??
    date.mmm.yy. %in% c('Jan-19','Jan-23'))

ggplot(data_long, aes(fill=date.mmm.yy., y=value, x=Category)) + 
  geom_bar(position="dodge", stat="identity", show.legend = FALSE) +
  geom_text(aes(label = date.mmm.yy.), position = position_dodge(width = 1), vjust = -0.5) +
  theme_classic() +
  labs(title = "Double Bar Graph",
       x = "Age",
       y = "All Population") +
  scale_x_discrete(labels = c("Under 16", "16 to 24", "25 to 44","45 to 64","65+"))
```

Figure 3 is a double bar graph comparing the total population of homeless people by age. We're able to see how the total numbers have changed from January 2019 to January 2023. There have been peaks and troughs throughout but this graph makes it simple to see things have changed.

```{r}
#| label: fig-4
#| fig-cap: "Percentage of Homeless Population in Toronto by Group"
#| echo: false
#| warning: false
#Table 1 : in a chunk of course in the quadro file !!
aux <- df %>% 
  filter(substr(date.mmm.yy., 1, 3)=='Jan' & year != 2018 &
           !population_group %in% c('All Population')) %>% 
  select(date.mmm.yy., population_group, population_group_percentage)
data_wide <- spread(aux, key = date.mmm.yy., value = population_group_percentage)
names(data_wide)[1] = 'Population Group'
kable(data_wide)
```

Table 1 provides data on the percentage of homeless individuals based on population group over the course of five years from January 2019 to January 2023. "Chronic" homelessness are individuals who have been homeless for a long period of time. The percentage of chronic homeless people have increased each year from 30.3% in January 2019  to 46.9% in January 2023.

@@

```{r}
#| label: fig-5
#| fig-cap: "Percentage of Homeless Population in Toronto by Group"
#| echo: false
#| warning: false
df$new_date = dmy(paste0("01-", df$date.mmm.yy.))
df$year = as.factor(year(df$new_date))
df$month = as.factor(month(df$new_date))
df$population_group = as.factor(df$population_group)
attach(df)
aux <- df %>% 
  filter(population_group == 'All Population') %>% 
  group_by(year, month) %>% 
  summarise(actively_homeless = sum(actively_homeless)) %>% 
  mutate(date = as.Date(paste0(year,'-',month,'-','01')))

ggplot(data = aux, aes(x = date, y = actively_homeless)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic() +
  labs(title = "Scatter Plot",
       x = "Date",
       y = "Actively Homeless")+
  scale_x_date(date_breaks = "12 months", date_labels = "%Y") 
```

```{r}
#| label: fig-6
#| fig-cap: "Percentage of Homeless Population in Toronto by Group"
#| echo: false
#| warning: false
df$new_date = dmy(paste0("01-", df$date.mmm.yy.))
df$year = as.factor(year(df$new_date))
df$month = as.factor(month(df$new_date))
df$population_group = as.factor(df$population_group)
attach(df)
aux = df[,c("newly_identified", "returned_from_housing", "returned_to_shelter",
            "moved_to_housing", "became_inactive", "new_date", "year", "month")] %>% 
  group_by(year, month) %>% 
  summarise(Newly_identified = sum(newly_identified),
            Returned_from_housing = sum(returned_from_housing),
            Returned_to_shelter = sum(returned_to_shelter),
            Moved_to_housing = sum(moved_to_housing),
            Became_inactive = sum(became_inactive)) %>% 
  mutate(date = as.Date(paste0(year,'-',month,'-','01')))

@@

ggplot(data = data_long, aes(x = date, y = value, color = Category)) + 
  geom_line(size = 2) +
  theme_classic() +
  labs(title = "Line Graph",
       x = "Date",
       y = "# of people")+
  scale_x_date(date_breaks = "12 months", date_labels = "%Y")+
  scale_color_discrete(name="Category",labels=c("Newly identified", 
              "Returned from housing","Returned to shelter",
              "Moved to housing", "Became inactive"))
```

$$
aux = df %>% 
  filter(population_group == 'All Population') %>% 
  select(year, month, actively_homeless) %>% 
  mutate(date = as.Date(paste0(year,'-',month,'-','01'))) %>% 
  mutate(new_date = seq(min(date),max(date), by = 'month'))

ggplot(aux, aes(x = new_date, y = actively_homeless)) +
  geom_line(color = "blue") +
  labs(title = "Sales over Time",
       x = "Time",
       y = "Actively homeless") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 year") +
  theme_bw()

$$ {#eq-bayes}

@eq-bayes seems useful, eh?

Here's a dumb example of how to use some references: In paper we run our analysis in `R` [@citeR]. We also use the `tidyverse` which was written by @thereferencecanbewhatever If we were interested in baseball data then @citeLahman could be useful.

We can use maths by including latex between dollar signs, for instance $\theta$.

Men accounted for 77 per cent of homeless deaths in 2022 and the age group with the highest percentage of deaths was people aged 40 to 59, at 34 per cent.

@@

#4.1	The Key Social Issue of Homelessness in Toronto is Solvable

@@

The data set in this paper shows that single males over the age of 25 years, have continued to increase at a steady pace in the homeless population, while youth under 24 years and families have declined. There is also a steep decline in refugees around 2020, a result of our borders being closed during Covid19. In 2021, the Indigenous population in the city of Toronto was 0.8% of the total population and yet, 11.4% of the active homeless population identified as Indigenous. @fig-1 Recognizes that certain groups, like Indigenous Peoples, face more discrimination and barriers to housing, the city is placing a greater focus on priority groups. This is why Indigenous Peoples was added as a sixth category of the population in 2021 and likely why there was a decrease in the percent of this group among the active homeless moving from 11.4% in 2021 to 9.5% in 2023 in table-1.

In the popular podcast “Solvable”, (@cite-1) a nonprofit organization called Community Solutions uses data-driven strategies with the goal of getting chronic homelessness to zero. The initiative recognizes that data is key to understanding how many people are homeless, what do they need, and then providing the necessary supports. Without data, the problem and the people are invisible. This initiative has seen chronic homelessness in communities in Texas and Illinois.

@@

#4.2	Which comes first? Oppression or homelessness?

People experiencing homelessness are some of the most vulnerable in our society. There is also a correlation between low income, race and social exclusion as explained in an article titled “Race, Homelessness and Health in Toronto” (@cite-2) which finds that people of colour not only suffer barriers due to racism, they have poorer health, and are overrepresented among the homeless population. The fact that Indigenous Peoples are also overrepresented in the homeless population (@fig-1) gives rise to the question about the relationship between historically oppressed groups and poverty. In the article “Racism, Homelessness, and the Social Exclusion of Indigenous Peoples in Canada” (@cite-3)  we learn that the impact of colonization and the removal of people from their land has escalated their chance of being unhoused or underhoused. As a result, it makes it hard to find employment, which means lower income and less access to good health care. Or is it the other way around? Poor health leads to lower income which makes it difficult to find housing. People who identify as LGBTQ and who are also tracked in the active housing data, are another group who are oppressed yet they are underrepresented with lower than average numbers when compared to other vulnerable groups. However, there is some evidence to show that this is a group who often choose to ‘stay in the closet’ to avoid discrimination and violence – good reasons to not report yourself as identifying as queer.

#4.3	The Supply and Demand for Affordable Housing 
When working to solve the housing problem, data is critical. So is the investment in more affordable housing options, rent controls and wrap around supports. The City of Toronto’s affordable housing website outlines various strategies to address homelessness. One is to incent developers to build more housing and rental units, another is the recent zoning law on laneway houses and another is to offer grants, like the Rapid Housing Initiative Grant, to organizations who can create and operate supportive housing. The results from graph number @fig-6   show that moving people into housing isn’t always successful. People moving back into the shelter system from housing is at @fig-6.. This could be an argument that building more housing is not a solution, however, the real issues as to why people are homeless in the first place are immensely complex and require holistic approaches and ongoing support to ensure long-term success.  Wrap around supports like how to budget for rent, addiction counselling or good mental health care, are necessary complements to the much needed expansion of affordable housing supplies. If a homeless person were to seek housing they'd be put on a waitlist. The average wait time for a subsidized bachelor apartment is 10 years. Scott's Mission a homeless shelter in downtown Toronto is fully booked every night and afraid to move out so they stay. Homeless shelters are a bandaid and not the solution. Shelters aren't safe spaces and the goal needs to be to house people. Supportive shelters can’t keep up with demand because of the intake of refugees, Toronto’s condo boom and the opioid crisis.

##4.4 Government Lack of Spending

@@

## Weaknesses and next steps

@@

From the data, there was no opportunity to understand the gap between active homelessness and the number of beds available. Capturing the percent of beds or available supportive housing units that were open in any given month or the percent of people who are actively seeking shelter but are turned away would be useful in determining supply and demand. 
Data on Indigenous Peoples was not collected until 2021 and it wasn’t clear if they were already included among the other groups in the prior years. Removing these numbers from one category and into another could make it look like a false decline (although the numbers were so small that it’s unlikely to affect a trend in a major way).

@@

@cite-1 :“Solvable Podcast - Ronald Young Jr.. Podcast.” Pushkin Industries, https://www.pushkin.fm/podcasts/solvable.
@cite2: "Frankish, C. James, et al. “Homelessness and Health in Canada: Research Lessons and Priorities - Canadian Journal of Public Health.” SpringerLink, Springer International Publishing, 1 Mar. 2005, https://link.springer.com/article/10.1007/BF03403700."
@cite-3: "“Definition of Indigenous Homelessness in Canada.” Definition of Indigenous Homelessness in Canada | The Homeless Hub, https://homelesshub.ca/IndigenousHomelessness."

>>>>

---
title: "Toronto's Homeless: Understanding the Causes and Seeking Solutions"
subtitle: "Homelessness "
author: 
  - Will Davidson
thanks: "Code and data are available at: https://github.com/willdavidson01/Toronto-Homeless"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "People suffering from homelessness in the City of Toronto have become an accepted part of the urban landscape.  With the shortage of affordable housing, high rents, poor economy and rising mental health concerns, it’s hard to imagine a viable solution to this critical social determinant of health.  Using OpenDataToronto, this paper analyzes data and identifies trends in different populations, gender, and age, as well as the current capacity to house people either in an emergency or permanent housing. It also challenges current assumptions and draws attention to the fact that today’s data can provide solutions to tomorrow’s needs."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Despite economic growth and increased development in the city, Toronto continues to struggle with the homelessness crisis. Actions and interventions such as public policies that support affordable housing in new high rise developments, grants to non-profits to provide wrap around services and programs for people experiencing homelessness and partnerships with hotels and recreation centers to offer temporary housing or out of the cold programs are all solid attempts to address the complex issues but have fallen short of solving the problem. Individuals have struggled through extreme weather conditions and overcrowded shelters, with waitlists of up to ten years for subsidized housing. The supportive shelters can't keep up with the demand because of the intake of refugees, Toronto's condo boom, and Toronto's fight against drug and alcohol abuse. Inadequate income, inaccessible housing, racism in employment, and housing markets area all factors contributing to homelessness. Homeless people are marginalized, and they're pushed to the side and deemed less important. The government needs to take urgent action to address the root causes of homelessness, as living on the streets puts individuals at risk of poor health and premature death. Should the government declare a state of emergency? Should more homeless shelters be built? Or is the root cause one of supply and demand for sustainable, affordable rents and access to homeownership for the average earning family. These critical questions need to be asked to move forward, and the first and perhaps the most important step required is to collect and analyze multiple data points that can drive decisions on how resources are best spent on the number of people who need housing.

The data used in the report is obtained from OpenDataToronto and the data set is called "Toronto Shelter System Flow.” It provides information on folks have been ‘actively homeless’ in the City of Toronto and accessing overnight services between 2019 and 2023. The data is organized using six different population groups. The dataset is used and explained to provide further analysis. Visualizations in graphs and tables are presented to illustrate the correlation between year, homelessness, category, age, and race. By data visualization and before constructing the model, we assumed the total population of actively homeless people over the past five years would be experiencing a steep rise. We present a multiple linear regression model to justify the relationship between homelessness and year. The interpretation of the final model and reasoning is shown in the latter section. A discussion is carried out and the the implications of the findings, including causes and solutions. We conclude with the following steps that need to be taken for a better future.

@@

@@

## Data Collection

The raw data was obtained from Opendatatoronto within the "Toronto System Shelter Flow". The data for the System Flow is captured through an information management system used to operate overnight shelters, 24-hour repsite sites, and warming centers funded by the City of Toronto. The data is calculated on the last day of each month and it considers all people who accessed overnight shelters. The analysis of the data was done through the R programming language [@citeR] with the 'tidyverse' [@citeTidy], 'dplyr' [@citeDplyr], and 'knitr' [@citeKnit] packages being used to clean the data. 'ggplot2' [@citeggplot2] was used to produce all the figures in this paper. The 'Lubridate' [@citeLub] package was also used to reformat dates."Toronto Shelter Flow" is the dataset used. (@citeDataset)

```{r}
#| label: fig-1
#| fig-cap: "Percentage of Homeless Population in Toronto by Group"
#| echo: false
#| warning: false
df$new_date = dmy(paste0("01-", df$date.mmm.yy.))
df$year = as.factor(year(df$new_date))
df$month = as.factor(month(df$new_date))
df$population_group = as.factor(df$population_group)
attach(df)
aux <- df %>% 
  filter(substr(date.mmm.yy., 1, 3)=='Jan' & year != 2018 &
           !population_group %in% c('All Population')) %>% 
  select(date.mmm.yy., population_group, population_group_percentage)
data_wide <- spread(aux, key = date.mmm.yy., value = population_group_percentage)
names(data_wide)[1] = 'Population Group'
kable(data_wide)
```

@fig-1 provides data on the percentage of homeless individuals based on population group over the course of five years from January 2019 to January 2023. "Chronic" homelessness are individuals who have been homeless for a long period of time. The percentage of chronic homeless people have increased each year from 30.3% in January 2019  to 46.9% in January 2023.

@@

## Data Visualization

@fig-2 shows the amount of actively homeless individuals in Toronto over the course of 2018 to 2022 with twelve scatter points per year. The graph shows the growth and decline ins with numbers on the overall rise. The fall in 2019 and 2020 can be attributed to the pandemic and the City of Toronto focused on helping those in need. They've since spent their money elsewhere and have let the numbers rise. The highest number at 11,000 and the lowest being in the 7,000 range are grossly high. They represnet a large part of the population.

```{r}
#| label: fig-2
#| fig-cap: "Percentage of Homeless Population in Toronto by Group"
#| echo: false
#| warning: false
df$new_date = dmy(paste0("01-", df$date.mmm.yy.))
df$year = as.factor(year(df$new_date))
df$month = as.factor(month(df$new_date))
df$population_group = as.factor(df$population_group)
attach(df)
aux <- df %>% 
  filter(population_group == 'All Population') %>% 
  group_by(year, month) %>% 
  summarise(actively_homeless = sum(actively_homeless)) %>% 
  mutate(date = as.Date(paste0(year,'-',month,'-','01')))

ggplot(data = aux, aes(x = date, y = actively_homeless)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic() +
  labs(title = "Percentage of Homeless Population in Toronto",
       x = "Date",
       y = "Actively Homeless")+
  scale_x_date(date_breaks = "12 months", date_labels = "%Y") 
```

I used the regression analysis approach.

Y = β0 + β1X1 + β2X2 + β3*X3 + ε

where:

Y is the number of homeless people in Toronto
X1 is age
X2 is race (encoded as a categorical variable with multiple binary indicators)
X3 is gender (encoded as a binary variable)
β0 is the intercept term
β1, β2, and β3 are the regression coefficients for each predictor variable
ε is the error term

I can use this to better understand the relationships between predicators and the number of homeless people in Toronto. An example of a positive coeffient could be as age increases the number of homeless people increase.

## Homeless Population Trends by Different Categories of the Population

```{r} 
#| label: fig-3
#| fig-cap: "Homeless Population Groups in Toronto From 2019 to 2023"
#| echo: false
#| warning: false
library(tidyverse)

ggplot(data = aux, aes(x = date, y = actively_homeless, color = population_group)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic() +
  labs(title = "Homeless Population Groups in Toronto From 2019 to 2023",
       x = "Date",
       y = "Actively Homeless")+
  scale_x_date(date_breaks = "12 months", date_labels = "%Y") 
```

@fig-3 represents a scatter plot that shows the trends of in Toronto's homelessness in the past five years. The Y-Axis is the number of current Torontoians who are actively homeless and the X-Axis are the twelve months in each of the past five years. The population groups are represented by individual data points.

## Toronto Homelessness based on Gender

```{r}
#| label: fig-4
#| fig-cap: "Homeless Population Based on Gender"
#| echo: false
#| warning: false
library(tidyverse)

ggplot(data = data_long, aes(x = date, y = value, color = Category)) + 
  geom_line(size = 2) +
  theme_classic() +
  labs(title = "Homeless Population Based on Gender",
       x = "Date",
       y = "Actively Homeless")+
  scale_x_date(date_breaks = "12 months", date_labels = "%Y") 
```

@fig-4 are multiple line graphs that compare the number of actively homeless people based on gender over the past five years.

## Toronto Homelessness by Age Group
```{r}
#| label: fig-5
#| fig-cap: "Homeless Population by Age Group in Toronto"
#| echo: false
#| warning: false
library(tidyverse)
#3rd graph
aux = df %>% 
  select(date.mmm.yy., ageunder16, age16.24, age25.44, age45.64, age65over) %>% 
  filter(population_group == 'All Population' & #or chronic ??
    date.mmm.yy. %in% c('Jan-19','Jan-23'))

ggplot(data_long, aes(fill=date.mmm.yy., y=value, x=Category)) + 
  geom_bar(position="dodge", stat="identity", show.legend = FALSE) +
  geom_text(aes(label = date.mmm.yy.), position = position_dodge(width = 1), vjust = -0.5) +
  theme_classic() +
  labs(title = "Homeless Population by Age Group in Toronto",
       x = "Age",
       y = "All Population") +
  scale_x_discrete(labels = c("Under 16", "16 to 24", "25 to 44","45 to 64","65+"))
```

@fig-5 is a double bar graph comparing the total population of homeless people by age. We're able to see how the total numbers have changed from January 2019 to January 2023. There have been peaks and troughs throughout but this graph makes it simple to see things have changed.

## Percentages of Toronto Homeless Groups
```{r}
#| label: fig-6
#| fig-cap: "Percentage of Homeless Population in Toronto by Group"
#| echo: false
#| warning: false
df$new_date = dmy(paste0("01-", df$date.mmm.yy.))
df$year = as.factor(year(df$new_date))
df$month = as.factor(month(df$new_date))
df$population_group = as.factor(df$population_group)
attach(df)
aux = df[,c("newly_identified", "returned_from_housing", "returned_to_shelter",
            "moved_to_housing", "became_inactive", "new_date", "year", "month")] %>% 
  group_by(year, month) %>% 
  summarise(Newly_identified = sum(newly_identified),
            Returned_from_housing = sum(returned_from_housing),
            Returned_to_shelter = sum(returned_to_shelter),
            Moved_to_housing = sum(moved_to_housing),
            Became_inactive = sum(became_inactive)) %>% 
  mutate(date = as.Date(paste0(year,'-',month,'-','01')))

@@

ggplot(data = data_long, aes(x = date, y = value, color = Category)) + 
  geom_line(size = 2) +
  theme_classic() +
  labs(title = "Percentage of Homeless Population in Toronto by Group",
       x = "Date",
       y = "# of people")+
  scale_x_date(date_breaks = "12 months", date_labels = "%Y")+
  scale_color_discrete(name="Category",labels=c("Newly identified", 
              "Returned from housing","Returned to shelter",
              "Moved to housing", "Became inactive"))
```

@fig-6 is a line graph that shows the trends for different categories of homeless people in Toronto. The Y-Axis shows the amount of people in each category.

The graph shows how the amount of newly identified homeless people as the highest with a peak of 4,000 people. Every category jumped in 2020 aside from those who return to housing the lowest steady at the bottom. The amount of people who are returning to the shelter are decreasing and this is due to high wait lists are the cause of them leaving and not being able to return. Many individuals are becoming inactive but we're not sure if it's because they're getting homes, passing away, or not in the shelter system and live off the grid.

@@

## The Key Social Issue of Homelessness in Toronto is Solvable

@@

The data set in this paper shows that single males over the age of 25 years, have continued to increase at a steady pace in the homeless population, while youth under 24 years and families have declined. There is also a steep decline in refugees around 2020, a result of our borders being closed during Covid19. In 2021, the Indigenous population in the city of Toronto was 0.8% of the total population and yet, 11.4% of the active homeless population identified as Indigenous. @fig-1 Recognizes that certain groups, like Indigenous Peoples, face more discrimination and barriers to housing, the city is placing a greater focus on priority groups. This is why Indigenous Peoples was added as a sixth category of the population in 2021 and likely why there was a decrease in the percent of this group among the active homeless moving from 11.4% in 2021 to 9.5% in 2023 in @fig-1.

In the popular podcast “Solvable”, [@cit1] a nonprofit organization called Community Solutions uses data-driven strategies with the goal of getting chronic homelessness to zero. The initiative recognizes that data is key to understanding how many people are homeless, what do they need, and then providing the necessary supports. Without data, the problem and the people are invisible. This initiative has seen chronic homelessness in communities in Texas and Illinois.

@@

##	Which comes first? Oppression or homelessness?

People experiencing homelessness are some of the most vulnerable in our society. There is also a correlation between low income, race and social exclusion as explained in an article titled “Race, Homelessness and Health in Toronto” [@cit2] which finds that people of colour not only suffer barriers due to racism, they have poorer health, and are overrepresented among the homeless population. The fact that Indigenous Peoples are also overrepresented in the homeless population (@fig-3) gives rise to the question about the relationship between historically oppressed groups and poverty. In the article “Racism, Homelessness, and the Social Exclusion of Indigenous Peoples in Canada” [@cit3]  we learn that the impact of colonization and the removal of people from their land has escalated their chance of being unhoused or underhoused. As a result, it makes it hard to find employment, which means lower income and less access to good health care. Or is it the other way around? Poor health leads to lower income which makes it difficult to find housing. People who identify as LGBTQ and who are also tracked in the active housing data, are another group who are oppressed yet they are underrepresented with lower than average numbers when compared to other vulnerable groups. However, there is some evidence to show that this is a group who often choose to ‘stay in the closet’ to avoid discrimination and violence – good reasons to not report yourself as identifying as queer.

##	The Supply and Demand for Affordable Housing

When working to solve the housing problem, data is critical. So is the investment in more affordable housing options, rent controls and wrap around supports. The City of Toronto’s affordable housing website outlines various strategies to address homelessness. One is to incent developers to build more housing and rental units, another is the recent zoning law on laneway houses and another is to offer grants, like the Rapid Housing Initiative Grant, to organizations who can create and operate supportive housing. The results from graph number @fig-6 show that moving people into housing isn’t always successful. People moving back into the shelter system from housing is at @fig-6. This could be an argument that building more housing is not a solution, however, the real issues as to why people are homeless in the first place are immensely complex and require holistic approaches and ongoing support to ensure long-term success.  Wrap around supports like how to budget for rent, addiction counselling or good mental health care, are necessary complements to the much needed expansion of affordable housing supplies. If a homeless person were to seek housing they'd be put on a waitlist. The average wait time for a subsidized bachelor apartment is 10 years. Scott's Mission a homeless shelter in downtown Toronto is fully booked every night and afraid to move out so they stay. Homeless shelters are a bandaid and not the solution. Shelters aren't safe spaces and the goal needs to be to house people. Supportive shelters can’t keep up with demand because of the intake of refugees, Toronto’s condo boom and the opioid crisis.

## Government Lack of Spending

@@

# Weaknesses and next steps

@@

From the data, there was no opportunity to understand the gap between active homelessness and the number of beds available. Capturing the percent of beds or available supportive housing units that were open in any given month or the percent of people who are actively seeking shelter but are turned away would be useful in determining supply and demand. Data on Indigenous Peoples was not collected until 2021 and it wasn’t clear if they were already included among the other groups in the prior years. Removing these numbers from one category and into another could make it look like a false decline (although the numbers were so small that it’s unlikely to affect a trend in a major way). The dataset only counts people who use overnight shelter systems making the real homelessness numbers much higher than reported. If the shelter doesn't use SMIS or funded by the government it's not included.

@@
